Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph\thanks{This work was funded by National Science Foundation award 1223057. Chris Whidden is a Simons Foundation Fellow of the Life Sciences Research Foundation.}

### 2015-11-04 curvature.tex
down to 15 pages for ANALCO proceedings

The automated computational analysis code can be found at \url{https://github.com/matsengrp/curvature}.

@@

The \emph{coarse Ricci-Ollivier curvature} of $x$ and $y$ is then defined as:
\vspace{-0.5em}
\begin{equation}
\curvature{m; x, y} := 1 - \frac{W_1(m_x, m_y)}{d(x, y)}.
\label{eq:curvatureDef}
\end{equation}
For the purposes of this paper, ``curvature'' without further specification will refer to \eqref{eq:curvatureDef}.
We will use $\curvature{x,y}$ to denote the curvature of the simple (uniform choice of neighbor) random walk, and use $\curvature{\MH; x,y}$ to indicate curvature with respect to the Metropolis-Hastings random walk sampling the uniform distribution (described in detail in Section~\ref{sec:random_walks}).
Positive curvature implies that the neighborhoods $m_x$ and $m_y$ are closer in transportation distance than point masses at $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are more distant than point masses at $x$ and $y$.
Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with the given random walk.
\begin{figure}[!h]
	\hspace*{\stretch{1}}
	\includegraphics[width=0.5\textwidth]{figs/neighborhood}
	\hspace*{\stretch{1}}
	\caption{The neighborhood of an $X$-tree $T$ with 4 leaves, showing connections between neighbors.}
	\label{fig:neighborhood}
\end{figure}
\vspace{0.1em}

@@

\bibliographystyle{siam}
\bibliography{curvature}

@@

\section{Supplementary Proofs}

@@

\degreemaxdelta*
\begin{proof}
	To prove (i), we simply note from Lemma~\ref{lem:degree_extremes} that the ladder tree achieves the minimum degree, and the balanced tree achieves the maximum degree:
	\begin{align*}
		\frac{\degree{T}}{\degree{S}} \ge\ &\frac{3n^2 - 13n + 14}{4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4(n-2)^2 - 2(n-2)} \\
		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 16n + 16 - 2(n-2)} \\
		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 18n + 20} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3, \\
	\end{align*}
	which is greater than 3/4 when $n \ge 3$.
    Similarly for (ii):
	\begin{align*}
		{\Delta}N &= \degree{S} - \degree{T} \\
		&\le\ (4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}) \\
		&\ \ \ \ \ \ \ \ - (3n^2 - 13n + 14) \\
		&\le\ (4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
		=\ &4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\
		&=\ n^2 - 5n + 6.
	\end{align*}
\end{proof}

\degreechange*
\begin{proof}
The set of permissible rSPR moves changes in four different ways due to the movement of $R$:
(i) subtrees that include nodes on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent node,
(ii) subtrees that include nodes on the path from $V$ to $L$ may no longer be moved into $R$ and its parent node,
(iii) $R$'s parent subtree may now be moved into $U$, and
(iv) $R$'s parent subtree may no longer be moved into $V$.
No additional moves are introduced or blocked by the original rSPR operation on $R$.

@@

Thus the difference in rSPR degree is $2ka - 2kb + 2(i-1) - 2(j-1)$ as claimed.
\end{proof}

\pairedneighbors*
\begin{proof}
By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $\overlap$ rSPR moves can be applied to $T$ and $S$ with the same source and target nodes.
For each such $(T',S')$ pair, we can move $R$ in either tree to obtain the other member of the pair.
\end{proof}

@@

	If the subtree of $k$ leaves is moved to the root, then we do not need to exclude the target branch from $k$ and $a$, gaining an additional leaf to balance the product $ka$ at the cost of increasing $j$.
	This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.
	Namely, we have $2(ka + 1 - j)$, where $j = n - k = a + 1$.
	Let ${\Delta}N = \degree{S} - \degree{T}$.
	If we move the additional leaf, we have:
	\begin{align*}
		{\Delta}N &\le 2\left(\ceilXXL{\frac{n}{2}}\floorV{\frac{n-2}{2}}  + 1 - \left(\floorV{\frac{n-2}{2}} + 1\right)\right) \\
		&= 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},
	\end{align*}
like before.
Similarly, if we do not move the additional leaf, we also have:
\begin{align*}
	{\Delta}N &\le 2 \left(\ceilV{\frac{n-2}{2}}\floorXXL{\frac{n}{2}} +1 -  \left(\ceilV{\frac{n-2}{2}} + 1\right)\right) \\
&= 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},
\end{align*}
proving (i).

@@

\curvaturedistancebound*
\begin{proof}
	Observe that the distance between neighbors of $T$ and $S$ is bounded between $\dspr{T,S} - 2$ and $\dspr{T,S} + 2$.
	For the curvature upper bound, we then have $\curvature{T,S} \le 1 - \frac{\dspr{T,S} - 2}{\dspr{T,S}} = \frac{2}{\dspr{T,S}}$.
	The lower bound follows similarly.
\end{proof}

\maxadjacentcurvature*
\begin{proof}
The maximum curvature between adjacent trees $T$ and $S$ occurs when their neighborhoods have maximum overlap and all other tree pairs are at distance 1.
By Lemma~\ref{lem:shared_neighbors} the maximum overlap is $6n-17$.
The amount of overlapping mass in the shared neighbors of $T$ and $S$ is thus $\frac{6n-17}{\max(\degree{T},\degree{S})}$.
The minimum mass transfer cost is thus $1 - \frac{6n-17}{\max(\degree{T},\degree{S})}$.
This is minimized when $\degree{T}=\degree{S}$ are as small as possible, that is $T,S$ are ladders and $\degree{T} = 3n^2 - 13n + 14$.

The maximum curvature is thus $1 - \frac{\degree{T} - (6n-17)}{\degree{T}} = \frac{6n - 17}{\degree{T}}= \frac{6n-17}{3n^2-13n+14}$.
\end{proof}

@@

\asymptotic*
\begin{proof}
	We first prove the lower bound in the uniform case, that is $\curvature{T,S} \le \ric{T,S}$.
	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy uniform case with parameter $p$.
	Recall that $\curvature{T,S} = \curvature[1]{T,S} = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\curvature[p]{T,S} / p = \left. \left(1 - \frac{W_1'(T,S)}{\dspr{T,S}}\right) \right/ p$.
	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p) \, \dspr{T,S},$$ by the simple mass transport program obtained by treating the mass at $T$ and $S$ as separate from that of the neighbors.
	Then:
	\begin{align*}
		\frac{\curvature[p]{T,S}}{p} &= \left. \left(1 - \frac{W_1'(T,S)}{\dspr{T,S}}\right) \right/ p \\
		&\ge \left. \left(1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}\right) \right/ p \\
		&= \frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p} \\
		&= 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
		&= \curvature{T,S}.
	\end{align*}

@@

\mhcurvature*
\begin{proof}
	We first prove the lower bound.
	By Lemma~\ref{lem:degree_max_delta_adjacent}, the quotient of degrees for two adjacent trees $\ge \frac{5}{6}$.
	Thus, the Hastings ratio is always $\ge \frac{5}{6}$.
	This implies that at most $\frac{1}{6}$ of the mass remains at tree $T$ in the mass distribution.
	Let $W_1(T,S)$ be the cost of an optimal mass transport for the uniform random walk from $T$ to $S$, and $W_1'(T,S)$ the cost for the MH random walk.
	Moreover, let $m_T(z)$ and $m_S(w)$ be the mass assigned for the uniform random walk and $m'_T(z)$ and $m'_S(w)$ be the mass assigned for the MH random walk, for each vertex $z \in N(T)$ and $w \in N(S)$.
	We construct an upper bound on $W_1'(T,S)$ by moving mass according to $W_1$ where possible, and moving the remainder either from $T$ to $S$, from $T$ to a neighbor of $S$, or from a neighbor of $T$ to $S$.
	That is, for each $W_1$ assignment $\xi(z,w)$, we send $\xi'(z,w) = \xi(z,w) \min\left(\frac{m'_T(z)}{m_T(z)}, \frac{m'_S(w)}{m_S(w)}\right)$ of the mass from $z$ to $w$.
	The remaining $\xi(z,w) - \xi'(z,w)$ of the mass is moved from $T$ to $S$, $T$ to $w$, and $z$ to $S$ in the respective proportions
	$\xi(z,w) \max\left(\frac{m'_T(z)}{m_T(z)}, \frac{m'_S(w)}{m_S(w)}\right) - \xi'(z,w)$,
	$\xi(z,w) \min\left(0, \frac{m'_T(z)}{m_T(z)} - \frac{m'_S(w)}{m_S(w)}\right)$, and,
	$\xi(z,w) \min\left(0, \frac{m'_S(w)}{m_S(w)} - \frac{m'_T(w)}{m_T(w)}\right)$.
	The maximum possible mass that is not moved according to $W_1$ is $\frac{1}{6}$.
	Moreover, the affected mass must be moved through at most two additional trees.
	Then, $W_1' \le W_1 + \frac{2}{6}$.
	We now have:
	\begin{align*}
		\curvature{\MH; T,S} & \ge 1 - \frac{W_1 + \frac{1}{3}}{\dspr{T,S}} \\
		& \ge \curvature{T,S} - \frac{1}{3\dspr{T,S}}.
	\end{align*}

@@

	We obtain the upper bounds similarly to the lower bounds, by observing that the affected at most $\frac{1}{6}$ of the mass may move through at most two fewer trees (i.e. directly between $T$ and $S$ rather than a pair of neighbors at distance $\dspr{T,S} + 2$ from each other).
	Again, this is at most one fewer tree when $\dspr{T,S}=1$.
\end{proof}

>>>>

The automated computational analysis code can be found at \url{https://github.com/matsengrp/curvature}.
Selected proofs of our theorems and lemmas can be found in the appendix.
Proofs omitted for space can be found in~\cite{curvature_arxiv}.

@@

The \emph{coarse Ricci-Ollivier curvature} of $x$ and $y$ is then defined as:
\vspace{-0.5em}
\begin{equation}
\curvature{m; x, y} := 1 - \frac{W_1(m_x, m_y)}{d(x, y)}.
\label{eq:curvatureDef}
\end{equation}
For the purposes of this paper, ``curvature'' without further specification will refer to \eqref{eq:curvatureDef}.
We will use $\curvature{x,y}$ to denote the curvature of the simple (uniform choice of neighbor) random walk, and use $\curvature{\MH; x,y}$ to indicate curvature with respect to the Metropolis-Hastings random walk sampling the uniform distribution (described in detail in Section~\ref{sec:random_walks}).
Positive curvature implies that the neighborhoods $m_x$ and $m_y$ are closer in transportation distance than point masses at $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are more distant than point masses at $x$ and $y$.
Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with a random walk.
\begin{figure}[!h]
	\hspace*{\stretch{1}}
	\includegraphics[width=0.43\textwidth]{figs/neighborhood}
	\hspace*{\stretch{1}}
	\caption{The neighborhood of an $X$-tree $T$ with 4 leaves, showing connections with $T$ and between neighbors.}
	\label{fig:neighborhood}
\end{figure}
\vspace{0.1em}

@@

\footnotesize
\bibliographystyle{siam}
\bibliography{curvature}
\normalsize

@@

\section{Selected Proofs}

@@

\cuttable{
\degreemaxdelta*
\begin{proof}
	To prove (i), we simply note from Lemma~\ref{lem:degree_extremes} that the ladder tree achieves the minimum degree, and the balanced tree achieves the maximum degree:
	\begin{align*}
		\frac{\degree{T}}{\degree{S}} \ge\ &\frac{3n^2 - 13n + 14}{4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4(n-2)^2 - 2(n-2)} \\
		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 16n + 16 - 2(n-2)} \\
		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 18n + 20} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3, \\
	\end{align*}
	which is greater than 3/4 when $n \ge 3$.
    Similarly for (ii):
	\begin{align*}
		{\Delta}N &= \degree{S} - \degree{T} \\
		&\le\ (4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}) \\
		&\ \ \ \ \ \ \ \ - (3n^2 - 13n + 14) \\
		&\le\ (4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
		=\ &4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\
		&=\ n^2 - 5n + 6.
	\end{align*}
\end{proof}
}

\cuttable{
\degreechange*
\begin{proof}
The set of permissible rSPR moves changes in four different ways due to the movement of $R$:
(i) subtrees that include nodes on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent node,
(ii) subtrees that include nodes on the path from $V$ to $L$ may no longer be moved into $R$ and its parent node,
(iii) $R$'s parent subtree may now be moved into $U$, and
(iv) $R$'s parent subtree may no longer be moved into $V$.
No additional moves are introduced or blocked by the original rSPR operation on $R$.

@@

Thus the difference in rSPR degree is $2ka - 2kb + 2(i-1) - 2(j-1)$ as claimed.
\end{proof}
}

\cuttable{
\pairedneighbors*
\begin{proof}
By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $\overlap$ rSPR moves can be applied to $T$ and $S$ with the same source and target nodes.
For each such $(T',S')$ pair, we can move $R$ in either tree to obtain the other member of the pair.
\end{proof}
}

@@

	If the subtree of $k$ leaves is moved to the root, then we do not need to exclude the target branch from $k$ and $a$, gaining an additional leaf to balance the product $ka$ at the cost of increasing $j$.
	This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.
	Namely, we have $2(ka + 1 - j)$, where $j = n - k = a + 1$.
	Let ${\Delta}N = \degree{S} - \degree{T}$.
	\cuttable[
		In both cases we have
		$${\Delta}N \le 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},$$
	]{
	If we move the additional leaf, we have:
	\begin{align*}
		{\Delta}N &\le 2\left(\ceilXXL{\frac{n}{2}}\floorV{\frac{n-2}{2}}  + 1 - \left(\floorV{\frac{n-2}{2}} + 1\right)\right) \\
		&= 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},
	\end{align*}
like before.
Similarly, if we do not move the additional leaf, we also have:
\begin{align*}
	{\Delta}N &\le 2 \left(\ceilV{\frac{n-2}{2}}\floorXXL{\frac{n}{2}} +1 -  \left(\ceilV{\frac{n-2}{2}} + 1\right)\right) \\
&= 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},
\end{align*}
}
proving (i).

@@

\cuttable{
\curvaturedistancebound*
\begin{proof}
	Observe that the distance between neighbors of $T$ and $S$ is bounded between $\dspr{T,S} - 2$ and $\dspr{T,S} + 2$.
	For the curvature upper bound, we then have $\curvature{T,S} \le 1 - \frac{\dspr{T,S} - 2}{\dspr{T,S}} = \frac{2}{\dspr{T,S}}$.
	The lower bound follows similarly.
\end{proof}
}

\cuttable{
\maxadjacentcurvature*
\begin{proof}
The maximum curvature between adjacent trees $T$ and $S$ occurs when their neighborhoods have maximum overlap and all other tree pairs are at distance 1.
By Lemma~\ref{lem:shared_neighbors} the maximum overlap is $6n-17$.
The amount of overlapping mass in the shared neighbors of $T$ and $S$ is thus $\frac{6n-17}{\max(\degree{T},\degree{S})}$.
The minimum mass transfer cost is thus $1 - \frac{6n-17}{\max(\degree{T},\degree{S})}$.
This is minimized when $\degree{T}=\degree{S}$ are as small as possible, that is $T,S$ are ladders and $\degree{T} = 3n^2 - 13n + 14$.

The maximum curvature is thus $1 - \frac{\degree{T} - (6n-17)}{\degree{T}} = \frac{6n - 17}{\degree{T}}= \frac{6n-17}{3n^2-13n+14}$.
\end{proof}
}

@@

\asymptotic*
\begin{proof}
	We first prove the lower bound in the uniform case, that is $\curvature{T,S} \le \ric{T,S}$.
	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy uniform case with parameter $p$.
	Recall that $\curvature{T,S} = \curvature[1]{T,S} = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\curvature[p]{T,S} / p = \left. \left(1 - \frac{W_1'(T,S)}{\dspr{T,S}}\right) \right/ p$.
	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p) \, \dspr{T,S},$$ by the simple mass transport program obtained by treating the mass at $T$ and $S$ as separate from that of the neighbors.
	Then:
	\begin{align*}
		\frac{\curvature[p]{T,S}}{p} &= \left. \left(1 - \frac{W_1'(T,S)}{\dspr{T,S}}\right) \right/ p \\
		&\ge \left. \left(1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}\right) \right/ p \\
		\cuttable{		&= \frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p} \\}
		&= 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
		&= \curvature{T,S}.
	\end{align*}

@@

\cuttable{
\mhcurvature*
\begin{proof}
	We first prove the lower bound.
	By Lemma~\ref{lem:degree_max_delta_adjacent}, the quotient of degrees for two adjacent trees $\ge \frac{5}{6}$.
	Thus, the Hastings ratio is always $\ge \frac{5}{6}$.
	This implies that at most $\frac{1}{6}$ of the mass remains at tree $T$ in the mass distribution.
	Let $W_1(T,S)$ be the cost of an optimal mass transport for the uniform random walk from $T$ to $S$, and $W_1'(T,S)$ the cost for the MH random walk.
	Moreover, let $m_T(z)$ and $m_S(w)$ be the mass assigned for the uniform random walk and $m'_T(z)$ and $m'_S(w)$ be the mass assigned for the MH random walk, for each vertex $z \in N(T)$ and $w \in N(S)$.
	We construct an upper bound on $W_1'(T,S)$ by moving mass according to $W_1$ where possible, and moving the remainder either from $T$ to $S$, from $T$ to a neighbor of $S$, or from a neighbor of $T$ to $S$.
	That is, for each $W_1$ assignment $\xi(z,w)$, we send $\xi'(z,w) = \xi(z,w) \min\left(\frac{m'_T(z)}{m_T(z)}, \frac{m'_S(w)}{m_S(w)}\right)$ of the mass from $z$ to $w$.
	The remaining $\xi(z,w) - \xi'(z,w)$ of the mass is moved from $T$ to $S$, $T$ to $w$, and $z$ to $S$ in the respective proportions
	$\xi(z,w) \max\left(\frac{m'_T(z)}{m_T(z)}, \frac{m'_S(w)}{m_S(w)}\right) - \xi'(z,w)$,
	$\xi(z,w) \min\left(0, \frac{m'_T(z)}{m_T(z)} - \frac{m'_S(w)}{m_S(w)}\right)$, and,
	$\xi(z,w) \min\left(0, \frac{m'_S(w)}{m_S(w)} - \frac{m'_T(w)}{m_T(w)}\right)$.
	The maximum possible mass that is not moved according to $W_1$ is $\frac{1}{6}$.
	Moreover, the affected mass must be moved through at most two additional trees.
	Then, $W_1' \le W_1 + \frac{2}{6}$.
	We now have:
	\begin{align*}
		\curvature{\MH; T,S} & \ge 1 - \frac{W_1 + \frac{1}{3}}{\dspr{T,S}} \\
		& \ge \curvature{T,S} - \frac{1}{3\dspr{T,S}}.
	\end{align*}

@@

	We obtain the upper bounds similarly to the lower bounds, by observing that the affected at most $\frac{1}{6}$ of the mass may move through at most two fewer trees (i.e. directly between $T$ and $S$ rather than a pair of neighbors at distance $\dspr{T,S} + 2$ from each other).
	Again, this is at most one fewer tree when $\dspr{T,S}=1$.
\end{proof}
}
